% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother


% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}



\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\usepackage{tcolorbox}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{bm}


\newtcolorbox{greybox}{
  colback=white,
  colframe=blue,
  coltext=black,
  boxsep=5pt,
  arc=4pt}
  
\newcommand{\sectionbreak}{\clearpage}

 
\newcommand{\ds}[4]{\sum_{{#1}=1}^{#3}\sum_{{#2}=1}^{#4}}
\newcommand{\us}[3]{\mathop{\sum\sum}_{1\leq{#2}<{#1}\leq{#3}}}

\newcommand{\ol}[1]{\overline{#1}}
\newcommand{\ul}[1]{\underline{#1}}

\newcommand{\amin}[1]{\mathop{\text{argmin}}_{#1}}
\newcommand{\amax}[1]{\mathop{\text{argmax}}_{#1}}

\newcommand{\ci}{\perp\!\!\!\perp}

\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\mb}[1]{\mathbb{#1}}
\newcommand{\mf}[1]{\mathfrak{#1}}

\newcommand{\eps}{\epsilon}
\newcommand{\lbd}{\lambda}
\newcommand{\alp}{\alpha}
\newcommand{\df}{=:}
\newcommand{\am}[1]{\mathop{\text{argmin}}_{#1}}
\newcommand{\ls}[2]{\mathop{\sum\sum}_{#1}^{#2}}
\newcommand{\ijs}{\mathop{\sum\sum}_{1\leq i<j\leq n}}
\newcommand{\jis}{\mathop{\sum\sum}_{1\leq j<i\leq n}}
\newcommand{\sij}{\sum_{i=1}^n\sum_{j=1}^n}
	
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Array Decomposition},
  pdfauthor={Jan de Leeuw},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{Array Decomposition}
\author{Jan de Leeuw}
\date{May 26, 2026}
\begin{document}
\maketitle
\begin{abstract}
TBD
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\setcounter{tocdepth}{3}
\tableofcontents
}

\sectionbreak

\textbf{Note:} This is a working manuscript which will be
expanded/updated frequently. All suggestions for improvement are
welcome. All qmd, tex, html, pdf, R, and C files are in the public
domain. Attribution will be appreciated, but is not required. The files
can be found at \url{https://github.com/deleeuw/arrays}

\sectionbreak

\section{Introduction}\label{introduction}

\section{Decomposition}\label{decomposition}

\subsection{Vectors}\label{vectors}

Suppose \(x\) is an element of \(n\)-dimensional Euclidean space, with
the usual inner product \(x'y\) and the squared norm
\(\|x\|^2:=\sqrt{x'x}\). We can decompose \(x\) into two orthogonal
parts, with the first part a vector with all elements equal to the mean
and the second part a vector with the deviations from the mean. We use a
\(\bullet\) replacing a subscript to indicate averaging over that
subscript. The elementwise decomposition is \begin{equation}
x_i=x_\bullet+(x_i-x_\bullet).\label{eq-onedimelem}
\end{equation}

Equation \eqref{eq-onedimelem} can be written as \begin{equation}
x=Px+(I-P)x,\label{eq-onedimproj}
\end{equation} where \(P\) projects on the subspace of all constant
vectors, i.e.~\(P:=n^{-1}ee'\) with \(e\) a vector with all elements
equal to one. Now \(P^2=P\) and thus \(P(I-P)=0\), which shows that
\(Px\) and \((I-P)x\) are orthogonal. Also, as Pythagoras used to say,
\begin{equation}
\|x\|^2=\|Px\|^2+\|I-Px\|^2.\label{eq-onedimpyth}
\end{equation} Note that equation \eqref{eq-onedimproj} is actually
correct for any square matrix \(P\), but the two components are
orthogonal if and only if \(P\) is a projector. Thus
\eqref{eq-onedimpyth} is also only true if \(P\) is a projector.

We can generalize equation \eqref{eq-onedimproj} to mutually orthogonal
subspaces \(S_1,\cdots,S_m\), with corresponding projectors
\(P_1,\cdots,P_m\). If \(P_1+\cdots+P_m=I\) we can have the orthogonal
decomposition \begin{equation}
x=P_1x+P_2x+\cdots P_mx,
\end{equation} and if \(P_1+\cdots+P_m\lesssim I\) in the Loewner
sense\footnote{The eigenvalues of $P_1+\cdots+P_m$ are less than or equal to one.}
then \begin{equation}
x=P_1x+P_2x+\cdots P_mx+(I-P_1-\cdots-P_m)x
\end{equation} Note that the ``residual'' \(I-P_1-\cdots-P_m\) is
orthogonal to all \(P_1,\cdots,P_m\).

Suppose \(U\) is an \(n\times n\) matrix with \(U'U=nI\) and with first
column equal to \(e\). Define \(\alpha:=n^{-1}U'x\) and
\(P_j:=n^{-1}u_ju_j'\). The \(P_j\) are projectors orthogonal to each
other. Then \begin{equation}
U\alpha=\sum_{j=1}^nP_jx=(\frac{1}{n}ee')x+(I-\frac{1}{n}ee')x=x_\bullet e+(x-x_\bullet).
\end{equation} Although \(\alpha\) depends on the choice of \(U\) the
product \(U\alpha\) is the same for all \(U\).

We can also derive the decomposition from a least squares projection
perspective. Suppose we want to minimize \(\|x-a\|^2\) over all \(a\) of
the form \(a=\alpha e\). The solution for \(\alpha\) is, of course,
\(Px=x_\bullet\). And we may want to minimize \(\|x-a\|^2\) over all
\(a\) satisfying \(e'a=0\), which gives the solution
\((I-P)x=x-x_\bullet\).

The projection formulation suggests some obvious further
generalizations. We can replace the one-dimensional subspace of all
constant vectors by an arbitrary subspace \(S\), and we can replace the
inner product \(x'y\) by an arbitrary inner product \(x'Ay\), with \(A\)
positive semi-definite. If \(H\) is an \(A\)-orthonormal basis for the
subspace \(S\), i.e.~if \(H'AH=I\), then \(Px=HH'Ax\) or \((x-Px)'Az=0\)
for all \(z\in S\).

In the next sections we analyse the same approaches we have used for
vectors to decompositions of linear spaces of matrices and arrays. This
is actually not as much of a generalization as it appears to be, because
matrices and arrays are elements of inner product spaces, just like
vectors are. What we will introduce are different notations and in most
cases different inner products.

\subsection{Matrices}\label{matrices}

Suppose \(X\) is an \(n\times m\) matrix, an element of the space
\(\mathbb{R}^{n\times m}\) with inner product
\(\langle X,Y\rangle:=\text{tr}\ X'Y\) and squared norm
\(\|X\|^2=\text{tr}\ X'X\). The elementwise decomposition we are
interested in is \begin{equation}
x_{ij}=x_{\bullet\bullet}+(x_{i\bullet}-x_{\bullet\bullet})+(x_{\bullet j}-x_{\bullet\bullet})+
(x_{ij}-x_{i\bullet}-x_{\bullet j}+x_{\bullet\bullet}).\label{eq-twodimelem}
\end{equation} Using ANOVA terminology we have the \emph{overall
effect}, the two \emph{main effects} for rows and columns, and the
\emph{first order interaction} of the rows and columns.

In the language of projectors we define \begin{subequations}
\begin{align}
P_r&:=n^{-1}ee',\\
P_c&:=m^{-1}ee',
\end{align}
\end{subequations} and with \(Q_r:=I-P_r\) and \(Q_c:=I-P_c\). Equation
\eqref{eq-twodimelem} becomes \begin{equation}
X=P_rXP_c+P_rXQ_c+Q_rXP_c+Q_rXQ_c.\label{eq-twodimproj}
\end{equation} Thus again we have four projectors, orthogonal to each
other, adding up to the identity.

In {de Leeuw} (\citeproc{ref-deleeuw_E_26e}{2026}) we generalize the
results in this section to the inner product
\(\langle X,Y\rangle:=\text{tr}\ RXCY'\), with \(R\) and \(C\) positive
semi-definite. Decomposition \eqref{eq-twodimproj} remains correct, with
\(P_r=ee'R/e'Re\) and \(P_c=Cee'/e'Ce\). Note that both projectors are
asymmetric. Equation \eqref{eq-twodimelem} also remains valid, if we
interpret the bullitt subscripts as weighted means.

To see more clearly how the vector case and the matrix case are
different we use the vec() operator to map \(\mathbb{R}^{n\times m}\)
into \(\mathbb{R}^{nm}\). Using the Kronecker product \(\otimes\) of
matrices we have the identity \begin{equation}
\text{tr}\ RXCY'=x'(C\otimes R)y,
\end{equation} with \(x:=\text{vec}(X)\) and \(y=\text{vec}(Y)\). Thus
our matrix results can be interpreted as a special case of the vector
results in \(\mathbb{R}^{nm}\), with inner product \(x'(C\otimes R)y\).

Now we switch to using orthogonal functions on a product space. This
part is inspired by an anonymous and unpublished manuscript that I read
around 1980 with a title that might have been ``Lois de Probabilité sur
un Ensemble Produit'' and perhaps the words ``Décomposition'' and/or
``Interaction'' were also in the title or subtitle. It was either by
Jean-Paul Benzécri or by one of his students or co-workers. I did have a
copy and I lost it, and Google can't find me a new one. The published
paper by (\citeproc{ref-bener_82}{\textbf{bener\_82?}}) discusses the
same materials, but in less detail.

Define \(U\) and \(V\) to be square matrices with first column equal to
\(e\), \(U'U=nI\) and \(V'V=mI\). Note this implies \(UU'=nI\) and
\(VV'=mI\), and thus \(UU'-ee'=nJ_n\) and \(VV'-ee'=mJ_m\). Define
\(\alpha_{ij}:=(nm)^{-1}u_i'Xv_j\). Then \begin{equation}
a_{ij}u_iv_j'=P_i^UXP^V_j,
\end{equation} where \(P_i^U:=n^{-1}u_iu_i'\) and
\(P_j^V:=\frac{1}{m}v_jv_j'\) project on the lines through the origin
and respectively \(u\) and \(v\). Using these projectors we have the
identity \begin{equation}
X=\sum_{i=1}^n\sum_{j=1}^m\alpha_{ij}u_iv_j'=\sum_{i=1}^n\sum_{j=1}^mP_i^UXP^V_j,
\end{equation} and \(\|X\|^2=\|A\|^2\), where \(A:=\{\alpha_{ij}\}\). If
\(i\neq k\) then \(P_i^UP_k^U=0\) and if \(j\neq l\) then
\(P_j^VP_l^V=0\). Moreover if \(I\) and \(K\) are disjoint subsets of
\(\{1,\cdots,n\}\) then \begin{equation}
\left(\sum_{i\in I}P_i^U\right)\left(\sum_{k\in K}P_k^U\right)=0.
\end{equation} Although \(P_i^U\) and \(P_j^V\) are projectors on
\(\mathbb{R}^n\) and \(\mathbb{R}^m\), they are only the building blocks
for the projectors on \(\mathbb{R}^{n\times m}\) that we use in this
section. Define \begin{equation}
\Pi_{ij}(X):=P_i^UXP_j^V
\end{equation} then \(\Pi_{ij}\) projects \(\mathbb{R}^{n\times m}\) on
the line through the origin and \(u_iv_j'\). Thus
\(\Pi_{ij}\circ\Pi_{ij}=\Pi_{ij}\) and \(\Pi_{ij}\circ\Pi_{kl}= 0\)
unless \(i=k\) and \(j=l\). Also \begin{equation}
\|X\|^2=\sum_{i=1}^n\sum_{j=1}^m\|\Pi_{ij}(X)\|^2
\end{equation}

\begin{subequations}
\begin{align}
\Pi_{11}(X)&=x_{\bullet\bullet}ee'.\\
\sum_{j\neq 1}^m\Pi_{1j}(X)&=e(x_{\bullet_j}-x_{\bullet\bullet})',\\
\sum_{i\neq 1}^m\Pi_{i1}(X)&=(x_{i\bullet}-x_{\bullet\bullet})e',\\
\sum_{i\neq 1}^n\sum_{j\neq 1}^m\Pi_{ij}(X)=J_nXJ_m&=(x_{ij}-x_{i\bullet}e'-ex_{j\bullet}'+x_{\bullet\bullet})
\end{align}
\end{subequations}

Although the \(\Pi_{ij}\) depend on the choice of \(U\) and \(V\) the
partial sums defining the main effects and the interaction are
independent of \(U\) and \(V\).

\subsection{Arrays}\label{arrays}

If \(X\) is an \(n\times m\times\ell\) array then the basic elementwise
decomposition we are interested in is \begin{align}
x_{ijk}&=x_{\bullet\bullet\bullet}+(x_{i\bullet\bullet}-x_{\bullet\bullet\bullet})+(x_{\bullet j \bullet}-x_{\bullet\bullet\bullet})+(x_{\bullet\bullet k}-x_{\bullet\bullet\bullet})\\
&+(x_{ij\bullet}-x_{i\bullet\bullet}-x_{\bullet j\bullet}+x_{\bullet\bullet\bullet})\\
&+(x_{i\bullet k}-x_{i\bullet\bullet}-x_{\bullet\bullet k}+x_{\bullet\bullet\bullet})\\
&+(x_{\bullet jk}-x_{\bullet j\bullet}-x_{\bullet\bullet k}+x_{\bullet\bullet\bullet})\\
&+(x_{ijk}
-x_{ij\bullet}-x_{i\bullet k}-x_{\bullet jk}+x_{i\bullet\bullet}+x_{\bullet j\bullet}+x_{\bullet\bullet k}-x_{\bullet\bullet\bullet})
\end{align}

\sectionbreak

\section{Correspondence Analysis}\label{correspondence-analysis}

Suppose \(F\) is a two-dimensional \(n\times m\) contingency table,
normalized to have grand total equal to one, with row marginals in a
diagonal matrix \(D\) and column marginals in a diagonal matrix \(E\).
Use the inner product \(\langle F,G\rangle=\text{tr}\ D^{-1}FE^{-1}G'\).
The projectors are \[
f_{ij}/d_ie_j=1+\sum_{s=1}^m u_{is}v_{js}
\]

\section{Log-linear Analysis}\label{log-linear-analysis}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\protect\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-deleeuw_E_26e}
{de Leeuw, Jan}. 2026. {``Matrix Decomposition with Kronecker
Weights.''}
\url{https://jansweb.netlify.app/publication/deleeuw-e-26-e/deleeuw-e-26-e.pdf}.

\end{CSLReferences}




\end{document}
